---
title: "scRNAseq_Hana_07_14_2023_seurat"
output: word_document
date: "2023-07-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

# Loading libraries and setting location paths
```{r}
# remove.packages('Matrix')
# install.packages("Matrix", repos="http://R-Forge.R-project.org")
library("Matrix")
library(Seurat)
# packageVersion("Seurat")
library(dplyr)
library(ggplot2)
setwd("Z:/ResearchHome/Groups/sapkogrp/projects/RNAseq/common/scRNAseq_Paul_cardiomyopathy/newvolume/analysis/combined_figures/")

rm(list = ls())

fig_path <- 'Z:/ResearchHome/Groups/sapkogrp/projects/RNAseq/common/scRNAseq_Paul_cardiomyopathy/newvolume/analysis/combined_figures/'

```


Below we've included a few convenience functions for saving images and reading/writing your Seurat object to disk. When you're working with larger datasets, it's usually a good idea to save your progress after computationally intensive steps so you can back track if you wish to do so.

```{r, echo=FALSE}
# Convenience functions
SaveFigure <- function(plots, name, type = "png", width, height, res){
  if(type == "png") {
    png(paste0(fig_path, name, ".", type),
      width = width, height = height, units = "in", res = 200)
  } else {
    pdf(paste0(fig_path, name, ".", type),
      width = width, height = height)
}
print(plots)
dev.off()
}

SaveObject <- function(object, name){
  saveRDS(object, paste0(data_path, name, ".RDS"))
}

ReadObject <- function(name){
  readRDS(paste0(data_path, name, ".RDS"))
}
```

# Reading in data
After reading in the data we'll perform basic filtering a on our expression matrix to remove low quality cells and uninformative genes. The parameter "min_genes" will keep cells that have at least 300 genes, and similarly, "min_cells" will keep genes that are expressed in at least 5 cells. Note: Seurat version 4.1 includes a convenience function to read Parse data from the DGE folder. If you would like to use this function, please skip the code block below and see the section "Reading in data with Seurat >= 4.1"

```{r, echo= TRUE}
DGE_folder <- "Z:/ResearchHome/Groups/sapkogrp/projects/RNAseq/common/scRNAseq_Paul_cardiomyopathy/newvolume/analysis/S_combined/all-well/DGE_filtered/"

# split-pipe versions older than 1.1.0 used "DGE.mtx"
mat <- readMM(paste0(DGE_folder, "DGE.mtx"))

cell_meta <- read.delim(paste0(DGE_folder, "cell_metadata.csv"),
                  stringsAsFactor = FALSE, sep = ",")
genes <- read.delim(paste0(DGE_folder, "all_genes.csv"),
                    stringsAsFactor = FALSE, sep = ",")

cell_meta$bc_wells <- make.unique(cell_meta$bc_wells, sep = "_dup")
rownames(cell_meta) <- cell_meta$bc_wells
genes$gene_name <- make.unique(genes$gene_name, sep = "_dup")

# Setting column and rownames to expression matrix
colnames(mat) <- genes$gene_name
rownames(mat) <- rownames(cell_meta)
mat_t <- t(mat)

# Remove empty rownames, if they exist
mat_t <- mat_t[(rownames(mat_t) != ""),]

# Seurat version 5 or greater uses "min.features" instead of "min.genes"
pbmc <- CreateSeuratObject(mat_t, min.features = 100, min.cells = 2, meta.data = cell_meta)
```

When we create our Seurat object the plate well numbers (column names in the expression matrix) from the experiment will automatically be assigned to the cell identity slot. In other words, the program assumes this how we want to initially classify our cells. In general, we would like to avoid this behavior so there isn't a bias towards a particular cell class when removing outliers.

```{r, echo=TRUE}
# Setting our initial cell class to a single type, this will changer after clustering. 
pbmc@meta.data$orig.ident <- factor(rep("pbmc", nrow(pbmc@meta.data)))
Idents(pbmc) <- pbmc@meta.data$orig.ident

SaveObject(pbmc, "seurat_obj_before_QC")
pbmc <- ReadObject("seurat_obj_before_QC")
```


# Cell quality control
In this step we'll perform cell quality control to prevent outlier cells from influencing downstream analyses. Cells that have unusually high transcript or gene counts are very likely to be multiplets, which is a term for two or more cells that have identical barcodes. Conversely, cells that have very low transcript or genes counts are a consequence of barcoding cells with damage membranes, or barcoding ambient RNA.

Filtering outliers can be accomplished by generating a violin plot for each cell feature and manually selecting the threshold we wish to remove cells. We'll also add another important cell feature that shows the percentage of mitochondrial genes expressed in each cell. Cells with high mitochondrial gene percentages should be removed, as they are likely to have lost cytoplasmic RNA from lysis or may have increased apoptosis (Luecken and Theis 2019)

```{r, echo=TRUE}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
plot <- VlnPlot(pbmc, pt.size = 0.10,
features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
SaveFigure(plot, "vln_QC", width = 12, height = 6)

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
SaveFigure((plot1 + plot2),"scatter_QC", width = 12, height = 6, res = 200)
```

```{r, echo=TRUE}
# Perform the filtering
pbmc <- subset(pbmc, subset = nFeature_RNA < 5000 & nCount_RNA < 20000 & percent.mt < 15)
```

# Normalizing the data
After removing unwanted cells from the dataset, the next step is to normalize the data. By default, we employ a global-scaling normalization method LogNormalize that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result. Normalized values are stored in pbmc[["RNA"]]@data.

```{r, echo=TRUE}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```

# Identification of highly variable features
We next calculate a subset of features that exhibit high cell-to-cell variation in the dataset (i.e, they are highly expressed in some cells, and lowly expressed in others). We and others have found that focusing on these genes in downstream analysis helps to highlight biological signal in single-cell datasets. Our procedure in Seurat3 is described in detail here, and improves on previous versions by directly modeling the mean-variance relationship inherent in single-cell data, and is implemented in the FindVariableFeatures function. By default, we return 2,000 features per dataset. These will be used in downstream analysis, like PCA.

```{r, echo=TRUE}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)
# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
SaveFigure((plot1 + plot2), "var_features", width = 12, height = 6)
```
# Scaling the data
Next, we apply a linear transformation ('scaling') that is a standard pre-processing step prior to dimensional reduction techniques like PCA. The ScaleData function:

Shifts the expression of each gene, so that the mean expression across cells is 0
Scales the expression of each gene, so that the variance across cells is 1
This step gives equal weight in downstream analyses, so that highly-expressed genes do not dominate
The results of this are stored in pbmc[["RNA"]]@scale.data

Warning! Generating a scaled data matrix can require a significant amount of RAM for large datasets and potentially crash your session. We therefore chose to use the default setting which only uses the variable gene set in the previous step. It's important to note that scaling the variable gene subset does not affect clustering; it only limits the number of genes available for plotting in the DoHeatMap function.

```{r, echo=TRUE}
pbmc <- ScaleData(pbmc)
```


# Perform linear dimensional reduction
Next we perform PCA on the scaled data. By default, only the previously determined variable features are used as input, but can be defined using features argument if you wish to choose a different subset.
Seurat provides several useful ways of visualizing both cells and features that define the PCA, including VizDimReduction, DimPlot, and DimHeatmap

```{r, echo=TRUE}
pbmc <- RunPCA(pbmc)
SaveObject(pbmc, "seurat_obj_after_PCA")

# pbmc <- ReadObject("seurat_obj_after_PCA")

# Examine and visualize PCA results a few different ways
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```


```{r, echo=TRUE}
plot <- VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
SaveFigure(plot, "viz_PCA_loadings", width = 10, height = 8)
```

```{r, echo=TRUE}
plot <- DimPlot(pbmc, reduction = "pca", group.by = "orig.ident")
SaveFigure(plot, "pc1_2_scatter", width = 8, height = 6)
```

In particular DimHeatmap allows for easy exploration of the primary sources of heterogeneity in a dataset, and can be useful when trying to decide which PCs to include for further downstream analyses. Both cells and features are ordered according to their PCA scores. Setting cells to a number plots the 'extreme' cells on both ends of the spectrum, which dramatically speeds plotting for large datasets. Though clearly a supervised analysis, we find this to be a valuable tool for exploring correlated feature sets.


```{r, echo=TRUE}
# Image doesn't save as png unless fast = FALSE
plot <- DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE, fast = FALSE)
SaveFigure(plot, "dim_heatmap1", width = 8, height = 6)
```
```{r, echo=TRUE}
plot <- DimHeatmap(pbmc, dims = 1:15, cells = 500, balanced = TRUE, fast = FALSE)
SaveFigure(plot, "dim_heatmap1_15", width = 12, height = 18)
```
# Determine the 'dimensionality' of the dataset
To overcome the extensive technical noise in any single feature for scRNA-seq data, Seurat clusters cells based on their PCA scores, with each PC essentially representing a 'metafeature' that combines information across a correlated feature set. The top principal components therefore represent a robust compression of the dataset. However, how many components should we choose to include? 10? 20? 100?

In Macosko et al, we implemented a resampling test inspired by the JackStraw procedure. We randomly permute a subset of the data (1% by default) and rerun PCA, constructing a 'null distribution' of feature scores, and repeat this procedure. We identify 'significant' PCs as those who have a strong enrichment of low p-value features.

```{r, echo=TRUE}
# NOTE: This process can take a long time for big datasets, comment out for expediency. More
# approximate techniques such as those implemented in ElbowPlot() can be used to reduce
# computation time
# pbmc <- JackStraw(pbmc, num.replicate = 100)
# pbmc <- ScoreJackStraw(pbmc, dims = 1:20)
```

An alternative heuristic method generates an 'Elbow plot': a ranking of principle components based on the percentage of variance explained by each one (ElbowPlot function). In this example, we can observe an 'elbow' around PC9-10, suggesting that the majority of true signal is captured in the first 10 PCs.

For this particular dataset, we chose not to use the jackstraw plot for PC selection due to the amount of time it takes for 65k cells, but instead opted to use the elbow plot where we selected PCs 1-30.

```{r, echo=TRUE}
plot <- ElbowPlot(pbmc,ndims = 50)
SaveFigure(plot, "PC_elbow_plot", width = 8, height = 10)
```
Identifying the true dimensionality of a dataset -- can be challenging/uncertain for the user. We therefore suggest these three approaches to consider. The first is more supervised, exploring PCs to determine relevant sources of heterogeneity, and could be used in conjunction with GSEA for example. The second implements a statistical test based on a random null model, but is time-consuming for large datasets, and may not return a clear PC cutoff. The third is a heuristic that is commonly used, and can be calculated instantly. In this example, all three approaches yielded similar results, but we might have been justified in choosing anything between PC 7-12 as a cutoff.

We chose 10 here, but encourage users to consider the following:

Dendritic cell and NK aficionados may recognize that genes strongly associated with PCs 12 and 13 define rare immune subsets (i.e. MZB1 is a marker for plasmacytoid DCs). However, these groups are so rare, they are difficult to distinguish from background noise for a dataset of this size without prior knowledge.
We encourage users to repeat downstream analyses with a different number of PCs (10, 15, or even 50!). As you will observe, the results often do not differ dramatically.
We advise users to err on the higher side when choosing this parameter. For example, performing downstream analyses with only 5 PCs does significantly and adversely affect results.

# Cluster the cells

Seurat v3 applies a graph-based clustering approach, building upon initial strategies in (Macosko et al). Importantly, the distance metric which drives the clustering analysis (based on previously identified PCs) remains the same. However, our approach to partitioning the cellular distance matrix into clusters has dramatically improved. Our approach was heavily inspired by recent manuscripts which applied graph-based clustering approaches to scRNA-seq data [SNN-Cliq, Xu and Su, Bioinformatics, 2015] and CyTOF data [PhenoGraph, Levine et al., Cell, 2015]. Briefly, these methods embed cells in a graph structure - for example a K-nearest neighbor (KNN) graph, with edges drawn between cells with similar feature expression patterns, and then attempt to partition this graph into highly interconnected 'quasi-cliques' or 'communities'.
 

As in PhenoGraph, we first construct a KNN graph based on the euclidean distance in PCA space, and refine the edge weights between any two cells based on the shared overlap in their local neighborhoods (Jaccard similarity). This step is performed using the FindNeighbors. function, and takes as input the previously defined dimensionality of the dataset (first 10 PCs).
 

To cluster the cells, we next apply modularity optimization techniques such as the Louvain algorithm (default) or SLM [SLM, Blondel et al., Journal of Statistical Mechanics], to iteratively group cells together, with the goal of optimizing the standard modularity function. The FindClusters function implements this procedure, and contains a resolution parameter that sets the 'granularity' of the downstream clustering, with increased values leading to a greater number of clusters. We find that setting this parameter between 0.4-1.2 typically returns good results for single-cell datasets of around 3K cells. Optimal resolution often increases for larger datasets. The clusters can be found using the Idents function.

```{r, echo=TRUE}
# install.packages("Matrix", reinstall = TRUE)
# library(Matrix)
# update.packages(ask = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:30)
pbmc <- FindClusters(pbmc, resolution = 0.30)
```

# Reorder clusters according to their similarity
This step isn't explicitly required, but can ease the burden of merging cell clusters (discussed further in the section "Merging clusters and labeling cell types") by reassigning each cluster by their position on a phylogenetic tree.

```{r, echo=TRUE}
pbmc <- BuildClusterTree(pbmc, reorder = TRUE, reorder.numeric = TRUE)
```


# Run non-linear dimensional reduction (UMAP/tSNE)
Seurat offers several non-linear dimensional reduction techniques, such as tSNE and UMAP, to visualize and explore these datasets. The goal of these algorithms is to learn the underlying manifold of the data in order to place similar cells together in low-dimensional space. Cells within the graph-based clusters determined above should co-localize on these dimension reduction plots. As input to the UMAP and tSNE, we suggest using the same PCs as input to the clustering analysis.

```{r, echo=TRUE}
pbmc <- RunUMAP(pbmc, dims = 1:30)
plot <- DimPlot(pbmc, reduction = "umap")
SaveFigure(plot, "umap_louvain_res_p3", width = 9, height = 8)
```

You can save the object at this point so that it can easily be loaded back in without having to rerun the computationally intensive steps performed above, or easily shared with collaborators.

```{r, echo=TRUE}
# SaveObject(pbmc, "seurat_obj_clustered")
pbmc <- ReadObject("seurat_obj_clustered")
```













