---
title: "RNAseq analysis for LVNC project-endocardial_cells"
output: word_document
date: "3-11-2024"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, eval=TRUE, echo= TRUE}
setwd("Z:/ResearchHome/Groups/sapkogrp/projects//RNAseq/common/LVNC_Hana_project/")
```

## Project description (Notes from Hana)
I have two sets of samples, on are cardiomyocytes (CM) and the other are
endocardial cells (EC) and would like to see the differences within each cell
type.

For each cell type I have 12 samples coming from 4 groups which have 3
replicates  (each replicate is one hiPSC line, so they are considered
biological replicates). I have one control group and three cardiomyopathy
groups. These four groups are : control, LVNC (left ventricular non-compaction
cardiomyopathy), HCM (hypertrophic cardiomyopathy) and DCM (dilated
cardiomyopathy).

Here are sample details:

Control: 19-3, 25-3, 26-3
Cardiomyopathy (DCM): GW10, GW53, GW168
Cardiomyopathy (HCM): GW129, GW167, GW169
Cardiomyopathy (LVNC): GW30, GW64, GW159


1. The first question is to see differences between control and cardiomyopathy
group. 

2. The second and more important question is that what is the differences
between LVNC and other cardiomyopathy groups. Specifically, we are interested
in the trabecular/compaction process and how the trabeculations is different
between LVNC and two other cardiomyopathy group.

# Loading libraries and setting location paths

```{r, eval=TRUE, echo= TRUE}

library(ggplot2)
library(ggfortify)
library(ggrepel)
library(dplyr)
library(DESeq2)
library(ggplot2)
library("knitr")
library("kableExtra")
```

# Preparing count data
```{r, eval=TRUE, echo= TRUE}
counts<-read.table(file='Z:/ResearchHome/Groups/sapkogrp/projects//RNAseq/common/LVNC_Hana_project/SAPKO-820324-STRANDED_RSEM_gene_count.2024-02-16_18-57-58.txt', header=TRUE)
# > dim(counts)
# [1] 60754    28


counts$geneSymbol[duplicated(counts$geneSymbol)] <- paste0(counts$geneSymbol, "_", counts$geneID)[duplicated(counts$geneSymbol)]

counts = dplyr::select(counts, -c(geneID, bioType, annotationLevel))
counts = counts[rowSums(counts[,-1])>20,]

rownames(counts)<-counts[,1]
counts2<-counts[,-1]

```



```{r, eval=TRUE, echo= TRUE}
## data preparaton
counts.matrix_ROBI<-as.matrix(counts2)
mode(counts.matrix_ROBI)<-"integer"

properties<-read.table('Z:/ResearchHome/Groups/sapkogrp/projects//RNAseq/common/LVNC_Hana_project/phenotype_LVNC_project.txt', sep = '\t', header=TRUE)
properties$Cell_Type=as.factor(properties$Cell_Type)
properties$Group=as.factor(properties$Group)
properties$Status=as.factor(properties$Status)
## only endocardial_cells
properties <- properties[properties$Cell_Type == "Endocardial_Cells",]
properties <- properties[, c("sample_ID", "Group", "Status")] # remove Cell_type
counts.matrix_ROBI <- counts.matrix_ROBI[,match(properties$sample_ID, colnames(counts.matrix_ROBI))]
colnames(counts.matrix_ROBI) == properties$sample_ID


## Create a DESeqDataSet (dds_ROBIS) from the count matrix and phenotype information.
dds_ROBIS<-DESeqDataSetFromMatrix(countData = counts.matrix_ROBI,colData = properties, design=~1)

## Perform variance-stabilizing transformation (vst) on the DESeqDataSet
rld_ROBIS<-vst(dds_ROBIS,blind = FALSE)
mat_ROBIS<-assay(rld_ROBIS)

## Performs principal component analysis (PCA) on the transformed data.
pcaData<-plotPCA(rld_ROBIS, intgroup=c("Group"), ntop=100, returnData=TRUE)
percentVar<-round(100 * attr(pcaData, "percentVar"))

## Plot PCA
g <- ggplot(pcaData, aes(PC1, PC2))
g <- g + geom_point(size = 5, aes(fill = Group, col = Group)) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(face = "bold", color = "black"),
    panel.border = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.title = element_text(face = "bold", color = "black"),  # Corrected line
    text = element_text(color = "black", face = "bold")
  )
# Save the plot as a TIFF file
tiff(
  file = "PCA_LVNC_Project_top_100_genes_endocardial_cells.tiff",
  width = 16,
  height = 14,
  units = 'cm',
  res = 300,
  compression = 'lzw'
)

# Print the plot
print(g)

# Close the TIFF device
dev.off()

## Select the top 500 variable genes based on row variances. Create a heatmap
## (phm) using pheatmap with clustering of rows and columns.
library(pheatmap)
properties2<-properties
rownames(properties2)<-properties[,1]
properties2<-subset(properties2, select=-c(sample_ID))
myMatrix<-SummarizedExperiment::assay(rld_ROBIS)
topVarGenes<-head(order(rowVars(myMatrix),decreasing = TRUE), 500)
myMatrixforheatmap<-myMatrix[topVarGenes,]
phm<-pheatmap(myMatrixforheatmap, scale="row", cluster_rows = TRUE, cluster_cols = TRUE, annotation_col = properties2, show_colnames = TRUE, show_rownames = FALSE)
tiff(file="Heatmap_LVNC_Project_top_500_genes_endocardial_cells.tiff",width = 18, height = 32, units = 'cm', res = 300, compression = 'lzw')
phm
dev.off()
```


## 1.The first question is to see differences between control and cardiomyopathy group. 

Here, we will conduct differential expression analysis between "Cardiomyopathy" and "Control" groups, filter and categorize significant genes, and create a volcano plot for visualization. The volcano plot highlights genes with significant changes in expression, facilitating the identification of differentially expressed genes.The design in this context includes the factors Status (indicating whether a sample is from the Cardiomyopathy or Control group).

```{r, eval=TRUE, echo= TRUE}
# Create a DESeqDataSet (ddset) from the count matrix (counts.matrix_ROBI) and
# phenotype information (properties2).
ddset = DESeqDataSetFromMatrix(
    countData = counts.matrix_ROBI, # raw counts
    colData = properties2, # phenotype data
    design = ~Status
)

deseq_object = DESeq(ddset)

# Extracts results from the DESeq analysis, focusing on the contrast between
# "Cardiomyopathy" and "Control" groups.
deseq_results = data.frame(results(
  deseq_object,
  contrast = c("Status", "Cardiomyopathy", "Control")
))
deseq_results = deseq_results[order(deseq_results$padj), ]
deseq_results_filtered = subset(deseq_results, pvalue < 0.05)
dim(deseq_results_filtered)
deseq_results$expression = ifelse(
  deseq_results$padj < 0.05 &
    (deseq_results$log2FoldChange) >= (1.5),
  "Up",
  ifelse(
    deseq_results$padj < 0.05 & (deseq_results$log2FoldChange) <= (-1.5),
    "Down",
    "Stable"
  )
)
deseq_results$expression = factor(deseq_results$expression, levels = c("Up", "Down", "Stable"))
# deseq_results = subset(deseq_results, abs(log2FoldChange)<5 & !is.na(padj))
deseq_results = subset(deseq_results,!is.na(padj))

write.table(deseq_results, file="LVNC_project_deseq_results_Cardiomyopathy_vs_Control_endocardial_cells.txt", sep="\t")

#Create Volcano Plot from DESEQ results
p = ggplot(data = deseq_results, aes(x = log2FoldChange,
                                     y = -log10(padj),
                                     colour = expression,
                                     label = rownames(deseq_results))) +
    geom_point(alpha = 0.4, size = 3.5) +
    scale_color_manual(values=c("red", "blue","grey")) +
    geom_vline(xintercept = c(-2, 2), lty=4, col="black", lwd=0.8) +
    geom_hline(yintercept = -log10(0.05), lty=4, col="black", lwd=0.8) +
    labs(x = expression(bold(log[2]~"fold-change")),
         y = expression(bold(-log[10]~"adjusted p-value"))) +
    theme_bw() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          text = element_text(size=16, colour = "black", face = "bold"),
          axis.text = element_text(colour = "black", face = "bold"),
          legend.title = element_blank()) +
    scale_x_continuous(breaks = seq(-10, 10,3), limits = c(-10, 10))

tiff(file="Volcano_plot_Cardiomyopathy_vs_Control_endocardial_cells.tiff",
     width = 18, height = 16, units = 'cm', res = 300, compression = 'lzw')
p
dev.off()
```

# Interpretation:
```{r, eval=TRUE, echo= F}
kable(head(deseq_results), format = "pandoc", caption = "DESeq2 Results (showing first 6 genes only)")
```

**baseMean**: Represents the average of normalized counts across all samples. For example, the gene MEG3 has an average expression level of 17.22.

**log2FoldChange**: Indicates the fold change in expression between the "Cardiomyopathy" and "Control" groups on a logarithmic scale. For instance, MEG3 has a log2 fold change of 21.98, suggesting a substantial increase in expression in the "Cardiomyopathy" group compared to the "Control" group.

**lfcSE (log2FoldChange Standard Error)**: Provides an estimate of the standard error associated with the log2 fold change. A smaller lfcSE indicates a more precise estimate.

**stat**: Represents the Wald statistic, a measure of how many standard errors the estimated log2 fold change is from zero. In this case, higher absolute values of the statistic suggest greater evidence against the null hypothesis.

**pvalue**: Indicates the statistical significance of the log2 fold change. A lower p-value suggests stronger evidence against the null hypothesis.

**padj (adjusted p-value)**: It is the p-value adjusted for false discovery rate (FDR).

**expression**: Categorizes genes based on their expression changes. "Up" indicates upregulation in "Cardiomyopathy," "Down" indicates downregulation, and "Stable" suggests no significant change.


## 2.The second and more important question is that what is the differences between LVNC and other cardiomyopathy groups. Specifically, we are interested in the trabecular/compaction process and how the trabeculations is different between LVNC and two other cardiomyopathy group.

Here, we will design our analysis to identify differentially expressed genes between LVNC and Other_Cardiomyopathy groups, considering the LVNC status as factors in the design formula.We specify the comparison of interest ("LVNC" and "Other_Cardiomyopathy") to compare the gene expression differences associated with the LVNC status.

```{r, eval=TRUE, echo= TRUE}
## remove controls from properties
properties3 <- properties2[properties2$Status != "Control", ]

# Create a new column in colData to indicate LVNC status
properties3$LVNC_status <-
  ifelse(properties3$Group=="LVNC",
         "LVNC",
         "Other_Cardiomyopathy")

counts.matrix_ROBI.3 <-
  counts.matrix_ROBI[, match(rownames(properties3), colnames(counts.matrix_ROBI))]

# Run DESeq2 analysis
ddset_lvnc_vs_others <- DESeqDataSetFromMatrix(
  countData = counts.matrix_ROBI.3,
  colData = properties3,
  design = ~ LVNC_status
)

deseq_object_lvnc_vs_others <- DESeq(ddset_lvnc_vs_others)
deseq_results_lvnc_vs_others <-
  data.frame(results(
    deseq_object_lvnc_vs_others,
    contrast = c("LVNC_status", "LVNC", "Other_Cardiomyopathy")
  ))
deseq_results_lvnc_vs_others <-
  deseq_results_lvnc_vs_others[order(deseq_results_lvnc_vs_others$padj), ]
deseq_results_filtered_lvnc_vs_others <-
  subset(deseq_results_lvnc_vs_others, pvalue < 0.05)
dim(deseq_results_filtered_lvnc_vs_others)


deseq_results_lvnc_vs_others$expression = ifelse(
  deseq_results_lvnc_vs_others$padj < 0.05 &
    (deseq_results_lvnc_vs_others$log2FoldChange) >=  (1.5),
  "Up",
  ifelse(
    deseq_results_lvnc_vs_others$padj < 0.05 &
      (deseq_results_lvnc_vs_others$log2FoldChange) <= (-1.5),
    "Down",
    "Stable"
  )
)


deseq_results_lvnc_vs_others$expression = factor(deseq_results_lvnc_vs_others$expression,
                                                 levels = c("Up", "Down", "Stable"))
deseq_results_lvnc_vs_others = subset(deseq_results_lvnc_vs_others,!is.na(padj))


# Write results to a file
write.table(deseq_results_lvnc_vs_others, file = "LVNC_project_deseq_results_LVNC_vs_Other_Cardiomyopathy_endocardial_cells.txt", sep = "\t")



# Create Volcano Plot
p_lvnc_vs_others <- ggplot(data = deseq_results_lvnc_vs_others, aes(x = log2FoldChange,
                                                                    y = -log10(padj),
                                                                    colour = expression,
                                                                    label = rownames(deseq_results_lvnc_vs_others))) +
  geom_point(alpha = 0.4, size = 3.5) +
  scale_color_manual(values = c("red", "blue", "grey")) +
  geom_vline(xintercept = c(-2, 2), lty = 4, col = "black", lwd = 0.8) +
  geom_hline(yintercept = -log10(0.05), lty = 4, col = "black", lwd = 0.8) +
  labs(x = expression(bold(log[2] ~ "fold-change")),
       y = expression(bold(-log[10] ~ "adjusted p-value"))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, colour = "black", face = "bold"),
        axis.text = element_text(colour = "black", face = "bold"),
        legend.title = element_blank()) +
  scale_x_continuous(breaks = seq(-10, 10, 3), limits = c(-10, 10))

# Save the plot as a TIFF file
tiff(file = "Volcano_plot_LVNC_vs_Other_Cardiomyopathy_endocardial_cells.tiff",
     width = 18, height = 16, units = 'cm', res = 300, compression = 'lzw')
print(p_lvnc_vs_others)
dev.off()
```

# Interpretation of results:
```{r, eval=TRUE, echo= F}
kable(head(deseq_results_lvnc_vs_others), format = "pandoc", caption = "LVNC and other cardiomyopathy (showing first 6 genesonly)")
```

**baseMean**: Represents the average of normalized counts across all samples. For example, the gene AC069277.1 has an average expression level of 12.35.

**log2FoldChange**: Indicates the fold change in expression between "LVNC" and "Other_Cardiomyopathy" groups on a logarithmic scale. For instance, AC069277.1 has a log2 fold change of -4.79, suggesting a substantial decrease in expression in the "LVNC" group compared to the "Other_Cardiomyopathy" group.

**lfcSE (log2FoldChange Standard Error)**: Provides an estimate of the standard error associated with the log2 fold change. A smaller lfcSE indicates a more precise estimate.

**stat**: Represents the Wald statistic, a measure of how many standard errors the estimated log2 fold change is from zero. In this case, higher absolute values of the statistic suggest greater evidence against the null hypothesis.

**pvalue**: Indicates the statistical significance of the log2 fold change. A lower p-value suggests stronger evidence against the null hypothesis.

**padj (adjusted p-value)**: It is the p-value adjusted for false discovery rate (FDR).

**expression**: Categorizes genes based on their expression changes. "Up" indicates upregulation in "Cardiomyopathy," "Down" indicates downregulation, and "Stable" suggests no significant change.


# Furthermore, we can compare LVNC vs DCM 
```{r, eval=TRUE, echo= TRUE}
# Run DESeq2 analysis
ddset_lvnc_vs_others <- DESeqDataSetFromMatrix(
  countData = counts.matrix_ROBI.3,
  colData = properties3,
  design = ~ Group
)

deseq_object_lvnc_vs_DCM <- DESeq(ddset_lvnc_vs_others)
deseq_results_lvnc_vs_DCM <-
  data.frame(results(
    deseq_object_lvnc_vs_DCM,
    contrast = c("Group", "LVNC", "DCM")
  ))
deseq_results_lvnc_vs_DCM <-
  deseq_results_lvnc_vs_DCM[order(deseq_results_lvnc_vs_DCM$padj), ]
deseq_results_filtered_lvnc_vs_DCM <-
  subset(deseq_results_lvnc_vs_DCM, pvalue < 0.05)
dim(deseq_results_filtered_lvnc_vs_DCM)


deseq_results_lvnc_vs_DCM$expression = ifelse(
  deseq_results_lvnc_vs_DCM$padj < 0.05 &
    (deseq_results_lvnc_vs_DCM$log2FoldChange) >=  (1.5),
  "Up",
  ifelse(
    deseq_results_lvnc_vs_DCM$padj < 0.05 &
      (deseq_results_lvnc_vs_DCM$log2FoldChange) <= (-1.5),
    "Down",
    "Stable"
  )
)


deseq_results_lvnc_vs_DCM$expression = factor(deseq_results_lvnc_vs_DCM$expression,
                                                 levels = c("Up", "Down", "Stable"))
deseq_results_lvnc_vs_DCM = subset(deseq_results_lvnc_vs_DCM,!is.na(padj))


# Write results to a file
write.table(deseq_results_lvnc_vs_DCM, file = "LVNC_project_deseq_results_LVNC_vs_DCM_endocardial_cells.txt", sep = "\t")



# Create Volcano Plot
p_lvnc_vs_DCM <- ggplot(data = deseq_results_lvnc_vs_DCM, aes(x = log2FoldChange,
                                                                    y = -log10(padj),
                                                                    colour = expression,
                                                                    label = rownames(deseq_results_lvnc_vs_DCM))) +
  geom_point(alpha = 0.4, size = 3.5) +
  scale_color_manual(values = c("red", "blue", "grey")) +
  geom_vline(xintercept = c(-2, 2), lty = 4, col = "black", lwd = 0.8) +
  geom_hline(yintercept = -log10(0.05), lty = 4, col = "black", lwd = 0.8) +
  labs(x = expression(bold(log[2] ~ "fold-change")),
       y = expression(bold(-log[10] ~ "adjusted p-value"))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, colour = "black", face = "bold"),
        axis.text = element_text(colour = "black", face = "bold"),
        legend.title = element_blank()) +
  scale_x_continuous(breaks = seq(-10, 10, 3), limits = c(-10, 10))

# Save the plot as a TIFF file
tiff(file = "Volcano_plot_LVNC_vs_DCM_endocardial_cells.tiff",
     width = 18, height = 16, units = 'cm', res = 300, compression = 'lzw')
print(p_lvnc_vs_DCM)
dev.off()
```

# Finally, LVNC vs HCM
```{r, eval=TRUE, echo= TRUE}
deseq_object_lvnc_vs_HCM <- DESeq(ddset_lvnc_vs_others)
deseq_results_lvnc_vs_HCM <-
  data.frame(results(
    deseq_object_lvnc_vs_HCM,
    contrast = c("Group", "LVNC", "HCM")
  ))
deseq_results_lvnc_vs_HCM <-
  deseq_results_lvnc_vs_HCM[order(deseq_results_lvnc_vs_HCM$padj), ]
deseq_results_filtered_lvnc_vs_HCM <-
  subset(deseq_results_lvnc_vs_HCM, pvalue < 0.05)
dim(deseq_results_filtered_lvnc_vs_HCM)


deseq_results_lvnc_vs_HCM$expression = ifelse(
  deseq_results_lvnc_vs_HCM$padj < 0.05 &
    (deseq_results_lvnc_vs_HCM$log2FoldChange) >=  (1.5),
  "Up",
  ifelse(
    deseq_results_lvnc_vs_HCM$padj < 0.05 &
      (deseq_results_lvnc_vs_HCM$log2FoldChange) <= (-1.5),
    "Down",
    "Stable"
  )
)


deseq_results_lvnc_vs_HCM$expression = factor(deseq_results_lvnc_vs_HCM$expression,
                                                 levels = c("Up", "Down", "Stable"))
deseq_results_lvnc_vs_HCM = subset(deseq_results_lvnc_vs_HCM,!is.na(padj))


# Write results to a file
write.table(deseq_results_lvnc_vs_HCM, file = "LVNC_project_deseq_results_LVNC_vs_HCM_endocardial_cells.txt", sep = "\t")



# Create Volcano Plot
p_lvnc_vs_HCM <- ggplot(data = deseq_results_lvnc_vs_HCM, aes(x = log2FoldChange,
                                                                    y = -log10(padj),
                                                                    colour = expression,
                                                                    label = rownames(deseq_results_lvnc_vs_HCM))) +
  geom_point(alpha = 0.4, size = 3.5) +
  scale_color_manual(values = c("red", "blue", "grey")) +
  geom_vline(xintercept = c(-2, 2), lty = 4, col = "black", lwd = 0.8) +
  geom_hline(yintercept = -log10(0.05), lty = 4, col = "black", lwd = 0.8) +
  labs(x = expression(bold(log[2] ~ "fold-change")),
       y = expression(bold(-log[10] ~ "adjusted p-value"))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size = 16, colour = "black", face = "bold"),
        axis.text = element_text(colour = "black", face = "bold"),
        legend.title = element_blank()) +
  scale_x_continuous(breaks = seq(-10, 10, 3), limits = c(-10, 10))

# Save the plot as a TIFF file
tiff(file = "Volcano_plot_LVNC_vs_HCM_endocardial_cells.tiff",
     width = 18, height = 16, units = 'cm', res = 300, compression = 'lzw')
print(p_lvnc_vs_HCM)
dev.off()
```